\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{float}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  urlcolor=blue!60!black,
  citecolor=blue!60!black
}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  frame=single,
  xleftmargin=0.5em,
  xrightmargin=0.5em
}

\title{\textbf{Solving OBELIX with Asymmetric PPO, Privileged Critics, and an Observation-Only Switch Policy}}
\author{CS780 Capstone Project}
\date{}

\begin{document}
\maketitle

\begin{abstract}
This report describes the final submitted controller for the CS780 capstone problem
\emph{OBELIX: the Warehouse Robot}. The task is a partially observable sequential
decision-making problem in which a mobile robot must find a box, make contact,
attach to it, and push it to the goal boundary while minimizing the damage caused by
large stuck penalties. The final submission is a two-branch controller: a wall-aware
actor trained with asymmetric proximal policy optimization (PPO) using a privileged
critic, and a no-wall difficulty-3 actor trained with high-throughput vectorized PPO.
At deployment, the policy uses only the legal 18-bit observation and memory derived
from past legal observations and past actions. No simulator state, reward, attachment
flag, coordinates, or hidden variable is read at inference time.

The final local evaluation target was difficulty 3 with horizon \(T=2000\), using the
weighted score
\[
  0.4 R_{\text{no-wall}} + 0.6 R_{\text{wall}}.
\]
On ten local seeds using \path{evaluate.py}, the selected submission obtained
\[
  R_{\text{no-wall}} = 1820.4,
  \qquad
  R_{\text{wall}} = -1142.8,
  \qquad
  0.4R_{\text{no-wall}} + 0.6R_{\text{wall}} = 42.48.
\]
The central lesson from this project is that OBELIX is not merely a search problem. It
is a conservative partially observable control problem with severe credit assignment,
large downside risk for incorrect forward motion, and different optimal behaviors in
wall and no-wall settings. The final design worked because it explicitly addressed all
three issues: partial observability, asymmetric training, and mode-dependent behavior.
\end{abstract}

\section{Introduction}

The OBELIX task is a reinforcement learning problem in the standard sense: an agent
interacts with an environment over time, chooses actions from observations, receives
scalar rewards, and aims to maximize long-run return. Unlike supervised learning,
correct actions are never provided directly. Instead, the controller must discover useful
behavior through trial-and-error interaction and must balance exploration against the
large cost of poor actions. This is especially important here because the task is not fully
observable, the box can blink or move before contact, and obstacle walls can create
states where a locally aggressive push policy is catastrophic.

From the assignment description, the robot must first \emph{find} the box, then
\emph{push} it after contact, and finally \emph{unwedge} it at the boundary. These are
human-interpretable phases, but they are not given directly to the agent. The policy
must infer them from the 18-bit observation stream. This makes the problem closer to a
POMDP than a fully observed MDP, and it explains why memory and history-dependent
features matter.

The rest of this report first formalizes the task as a POMDP, then explains the reward
geometry and legal observation space, then presents the final asymmetric-PPO switch
architecture, the training setup, the alternatives that were tested and rejected, and the
final empirical results together with their interpretation.

\section{Formal Problem Definition}

\subsection{Environment Description}

OBELIX is a bounded two-dimensional robot-control task. A circular robot moves in a
square arena and must push a square box to the goal boundary. In the hardest setting
used for the final submission, the box can blink on and off and can move before
attachment. In the wall-obstacle version, a wall with a small opening can appear,
forcing the robot to choose approach angles and pushing directions carefully.

The action set is discrete:
\[
  \mathcal{A} = \{\texttt{L45},\texttt{L22},\texttt{FW},\texttt{R22},\texttt{R45}\}.
\]
The turn actions rotate the robot by \(45^\circ\) or \(22.5^\circ\). The \texttt{FW} action
attempts a forward move of 5 pixels. When the box is attached, forward motion attempts
to translate both the robot and the box.

Episodes terminate either when the box is successfully removed or when the horizon
\(T\) is reached. For the final evaluation, \(T=2000\).

\subsection{POMDP Formulation}

The interaction is naturally written as a partially observable Markov decision process
(POMDP)
\[
  \mathcal{P}=(\mathcal{S},\mathcal{A},\Omega,P,O,r,\rho_0,\gamma,H).
\]
Each component is defined below.

\paragraph{Latent state space \(\mathcal{S}\).}
The simulator state
\(s_t\in\mathcal{S}\) contains all variables required to make the dynamics Markov. In
this task that includes at least the robot pose, robot heading, box pose, box velocity,
box visibility/blinking state, attachment state, arena boundary geometry, wall geometry
when obstacles are enabled, and the internal variables required by the simulator to
resolve collisions and termination.

\paragraph{Action space \(\mathcal{A}\).}
The agent chooses one of the five legal discrete actions at each time step.

\paragraph{Observation space \(\Omega\).}
The deployed agent does not observe \(s_t\) directly. Instead it receives a binary
observation
\[
  o_t\in\Omega=\{0,1\}^{18}.
\]
This is the legal sensor interface exposed by the environment.

\paragraph{Transition kernel \(P\).}
The transition model
\[
  P(s_{t+1}\mid s_t,a_t)
\]
contains the motion of the robot, the pre-attachment motion of the box, the blinking
process, collision resolution, wall interactions, and the state update after contact or
goal completion.

\paragraph{Observation model \(O\).}
The observation process is
\[
  O(o_t\mid s_t).
\]
This maps the latent simulator state to the legal 18-bit sensor vector. Because many
different latent states can produce the same observation, the agent must infer task phase
and hidden structure from observation histories rather than from a single frame.

\paragraph{Reward function \(r\).}
The scalar reward is generated from the current transition. It includes a per-step cost,
large stuck penalties, one-time sensor rewards, an attachment bonus, and a large success
bonus when the box reaches the boundary.

\paragraph{Initial-state distribution \(\rho_0\).}
At the beginning of each episode the latent state is sampled from a distribution over
robot pose, box pose, box velocity, blinking mode, and wall configuration. This matters
because the policy must work over many randomized starts rather than a single fixed
layout.

\paragraph{Discount factor \(\gamma\).}
The PPO training objective uses discounted return with \(\gamma\in(0,1)\). For the final
wall branch, \(\gamma=0.995\). This favors high reward sooner rather than later and is
consistent with the capstone objective of solving the task quickly and safely.

\paragraph{Horizon \(H\).}
The environment is episodic with finite horizon \(H=T\). The final evaluation used
\(T=2000\).

\subsection{History-Dependent Policy}

Because the observation is partial, the optimal action generally depends on history, not
just the current observation. Formally, the deployed controller is not a memoryless map
from \(o_t\) to \(a_t\); it is a history-dependent policy
\[
  \pi(a_t\mid h_t),
  \qquad
  h_t=(o_0,a_0,o_1,a_1,\dots,o_t).
\]
In the actual implementation, the full history is compressed into finite legal memory
features built from observation stacks, action history, masks, and counters. This is the
correct way to make a POMDP manageable without violating the rules.

\subsection{Task Phases as Latent Modes}

Although the simulator does not expose symbolic stage labels, the task can be understood
as three latent behavioral modes:
\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{Search / find:} the robot must rotate and move to bring the box into the
  sensor sectors.
  \item \textbf{Approach / attach:} the robot must use local sensor geometry to make
  contact from a favorable side.
  \item \textbf{Push / unwedge:} after contact, the robot must drive the box to the goal
  boundary while avoiding wall traps and repeated stuck penalties.
\end{enumerate}
These modes are not given to the policy. They must be inferred from the observation
stream. This is one of the main reasons plain reactive policies were unreliable.

\section{Legal Observation Space}

The raw observation is an 18-bit binary vector
\[
  o_t=[b_0,\ldots,b_{15},b_{\text{IR}},b_{\text{stuck}}].
\]
The first 16 bits are sonar-like range bits. They are naturally grouped by direction:
\[
\begin{aligned}
  \text{left}  &: \quad b_0,\ldots,b_3,\\
  \text{front} &: \quad b_4,\ldots,b_{11},\\
  \text{right} &: \quad b_{12},\ldots,b_{15}.
\end{aligned}
\]
The final two bits are the infrared sensor and the stuck flag:
\[
  b_{\text{IR}}=b_{16},
  \qquad
  b_{\text{stuck}}=b_{17}.
\]
The infrared bit indicates that the grey box is directly ahead; the stuck bit indicates
that the robot is jammed against a wall or boundary.

The submitted policy never reads the simulator state, reward, done flag, box
coordinates, robot coordinates, attachment flag, or any other hidden variable. All
inference-time memory is constructed only from legal observations and the policy's own
past actions.

\section{Reward Dynamics and Credit Assignment}

The reward function is unusually conservative. A stationary policy that avoids all useful
contact still incurs the step cost, so over a horizon of \(T\) steps it can score roughly
\(-T\). At \(T=2000\), this means a passive agent is already close to \(-2000\). A
policy that repeatedly drives into a wall can be far worse because the stuck penalty is
large and can occur many times.

The per-step reward implemented by \path{obelix.py} is summarized as
\[
  r_t =
  -1
  - 200\,\mathbf{1}\{b_{\text{stuck}}=1\}
  + r_{\text{sensor-once}}
  + r_{\text{attach}}
  + r_{\text{success}}.
\]
The one-time sensor rewards are
\[
\begin{array}{ll}
\text{left/right sonar bits} & +1 \text{ per bit, once per episode},\\
\text{front far bits}       & +2 \text{ per bit, once per episode},\\
\text{front near bits}      & +3 \text{ per bit, once per episode},\\
\text{IR bit}               & +5 \text{ once per episode}.
\end{array}
\]
First attachment gives \(+100\), followed by the normal push-state step cost. Pushing
the attached box to a boundary gives a success bonus of \(+2000\).

This reward geometry creates a hard credit-assignment problem:
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Search is necessary} because the box may be invisible, moving, or outside
  the sensing region.
  \item \textbf{Risk is asymmetric:} a single useful forward move may help only slightly,
  but a single bad forward move can produce a \(-201\) transition.
  \item \textbf{Attachment is not enough:} contact from the wrong side often leads to a
  jammed push and repeated stuck penalties.
  \item \textbf{Myopic safety is misleading:} a policy that mostly rotates or hesitates can
  avoid catastrophic penalties, but still fail the task and accumulate a large negative
  return from the step cost.
\end{itemize}

This explains why many early agents looked active in rollouts but still scored poorly.
The policy had to be exploratory enough to find the box, structured enough to infer task
phase from partial observations, and conservative enough to avoid repeated collision
loops.

\section{Observation Engineering}

\subsection{No-Wall PPO Encoding}

For the no-wall branch, the best checkpoint used a compact encoded observation of
dimension 32. It concatenates the raw 18-bit observation with 14 derived features:
\[
  \phi_{\text{nw}}(o_t)=[o_t,d_t]\in\mathbb{R}^{32},
\]
where \(d_t\) contains
\[
\begin{array}{ll}
\text{directional hit flags} & \mathbf{1}\{\text{left hit}\},\; \mathbf{1}\{\text{front hit}\},\; \mathbf{1}\{\text{right hit}\},\\
\text{terminal bits} & b_{\text{IR}},\; b_{\text{stuck}},\; \mathbf{1}\{\text{blind}\},\\
\text{counts} & n_{\text{left}},\; n_{\text{front}},\; n_{\text{right}},\; n_{\text{front-far}},\; n_{\text{front-near}},\\
\text{differences} & n_{\text{left}}-n_{\text{right}},\; n_{\text{right}}-n_{\text{left}},\;
  n_{\text{front}}-\frac{1}{2}(n_{\text{left}}+n_{\text{right}}).
\end{array}
\]
This representation made the relative salience of front versus side hits directly visible
to a small multilayer perceptron and was sufficient in the no-wall difficulty-3 setting,
where the main challenge was search rather than fine obstacle negotiation.

\subsection{Wall Actor Memory Features}

In the wall setting, a single observation was usually insufficient. The deployed wall
actor uses the legal feature vector
\[
  x_t=
  [
    o_{t-11:t},
    \operatorname{onehot}(a_{t-6:t-1}),
    m_t,
    c_t
  ],
\]
where \(o_{t-11:t}\) is a 12-frame observation stack, the action history stores the last 6
actions, \(m_t\in\{0,1\}^{17}\) is a sensor-seen mask over bits \(0\ldots16\), and
\(c_t\in\mathbb{R}^8\) contains normalized counters:
\[
  c_t=
  [
  \text{blind streak},
  \text{stuck streak},
  \text{last-seen side},
  \text{contact memory},
  \text{step fraction},
  \text{repeated-observation streak},
  \text{blind-turn streak},
  \text{stuck memory}
  ].
\]
For the final wall checkpoint,
\[
  \dim(x_t)=12\cdot18 + 6\cdot5 + 17 + 8 = 271.
\]
All of these features are legal because they are deterministic functions of current and
past legal observations together with the actions already emitted by the policy.

\section{Learning Method: Why PPO and Why Asymmetric Training}

\subsection{Why a Policy-Gradient Method Was Used}

Several families of methods were plausible from the course material: tabular methods,
Monte Carlo control, TD methods such as Q-learning, and policy-gradient methods. In
practice, the deployed solution used PPO because the task combines partial observability,
large downside risk, reward shaping during training, and function approximation.

A tabular method is not suitable here because the latent state is continuous and hidden.
Even the legal observation history is too large to tabulate. Pure Monte Carlo updates are
also unattractive because they use the full return only at episode end, which causes high
variance. This is particularly problematic here because episodes are long and successful
trajectories are sparse. Temporal-difference methods reduce variance through
bootstrapping, but off-policy value-based learning was unstable in this project because
the observation aliasing and heavy penalties produced noisy Q-targets. PPO offered a
more stable compromise: direct policy optimization with a learned critic and generalized
advantage estimation (GAE), which is itself a bias-variance tradeoff.

\subsection{Asymmetric PPO}

The main training idea was to use asymmetric actor-critic learning. The deployed actor
receives only legal features \(x_t\), while the critic receives both \(x_t\) and privileged
simulator features \(p_t\):
\[
  \pi_\theta(a_t\mid x_t),
  \qquad
  V_\psi(x_t,p_t).
\]
This improves the value estimate during training without leaking privileged information
into the final submission.

The privileged vector has dimension 34. It includes normalized robot position,
\(\sin/\cos\) robot heading, box position, relative robot-box vector, relative distance,
relative heading, push-active flag, visibility flag, stuck flag, step fraction, wall flag,
difficulty, box speed, box velocity, boundary margins, goal distance, wall-gap geometry,
and terms indicating which side of the wall the robot and box occupy. These inputs are
used only by the training critic.

\subsection{Objective Functions}

The PPO actor objective uses the clipped surrogate
\[
  L_{\pi}(\theta)=
  \mathbb{E}_t\left[
    \min\left(
      \rho_t(\theta)\hat{A}_t,
      \operatorname{clip}(\rho_t(\theta),1-\epsilon,1+\epsilon)\hat{A}_t
    \right)
  \right],
\]
where
\[
  \rho_t(\theta)=
  \frac{\pi_\theta(a_t\mid x_t)}{\pi_{\theta_{\text{old}}}(a_t\mid x_t)}.
\]
Advantages were computed with generalized advantage estimation:
\[
  \hat{A}_t=
  \sum_{l=0}^{\infty}(\gamma\lambda)^l\delta_{t+l},
  \qquad
  \delta_t=r_t + \gamma V_\psi(x_{t+1},p_{t+1}) - V_\psi(x_t,p_t).
\]
The critic was trained with a clipped value loss and the actor with entropy
regularization. The entropy term mattered in practice because otherwise the policy often
collapsed into low-mobility turn-heavy behavior or overly aggressive repeated forward
pushing.

\subsection{Bias-Variance Interpretation}

This training setup can be interpreted using the course discussion of MC versus TD.
Pure Monte Carlo uses the full return and is unbiased but high variance; pure TD
bootstraps from the critic and is lower variance but biased. PPO with GAE lies between
these extremes. In this project that tradeoff was useful: the task needed lower-variance
updates than pure return targets, but still required enough long-horizon signal to learn
that early search and correct attachment geometry affect the final success bonus.

\section{Final Model Architecture}

\subsection{Wall Branch}

At inference time, the wall branch is an actor-only MLP:
\[
  271 \rightarrow 512 \rightarrow 256 \rightarrow 128 \rightarrow 5.
\]
Hidden activations are \(\tanh\), and the output is a 5-dimensional logit vector over the
five discrete actions. The checkpoint in the final submission contains both the actor
weights and the full asymmetric training checkpoint for reproducibility, but only the
actor backbone and policy head are loaded during deployment.

The wall branch was trained with \path{assym_ppo/train_asym_ppo.py}. The checkpoint
metadata for the deployed wall actor records:
\[
\begin{array}{ll}
\text{number of vectorized environments} & 512,\\
\text{rollout length} & 128,\\
\text{training steps} & 1{,}000{,}000,\\
\text{actor hidden sizes} & [512,256,128],\\
\text{critic hidden sizes} & [1024,512,256],\\
\gamma & 0.995,\\
\lambda_{\text{GAE}} & 0.95,\\
\text{learning rate} & 1.5\times 10^{-4},\\
\text{entropy coefficient} & 0.002,\\
\text{warm-start expert} & \texttt{eval300},\\
\text{warm-start episodes} & 20.
\end{array}
\]
Reward shaping during this phase included visibility and IR bonuses, an enlarged stuck
penalty, approach progress, alignment progress, and push progress. The policy was still
screened using the environment reward, not only the shaped training reward.

\subsection{No-Wall Branch}

The no-wall branch is a standard PPO actor-critic checkpoint deployed as an actor MLP:
\[
  32 \rightarrow 128 \rightarrow 64 \rightarrow 5.
\]
Its checkpoint in the final bundle is \path{ppo_lab/nowall_d3_ppo_v3.pth}. It was
trained using \path{PPO/train_ppo.py} with the CUDA vectorized environment:
\[
\begin{array}{ll}
\text{environment backend} & \texttt{torch\_vec},\\
\text{environment device} & \texttt{cuda},\\
\text{number of environments} & 256,\\
\text{rollout length} & 128,\\
\text{total environment steps} & 4{,}194{,}304,\\
\text{difficulty} & 3,\\
\text{wall obstacles} & \texttt{False},\\
\text{max steps} & 2000,\\
\text{learning rate} & 2\times10^{-4},\\
\text{entropy coefficient} & 0.005,\\
\text{update epochs} & 6,\\
\text{initial checkpoint} & \texttt{ppo\_lab/nowall\_d3\_ppo\_v1.pth}.
\end{array}
\]
This branch was retained because it consistently produced much stronger no-wall returns
than the recurrent and hand-written alternatives.

\section{Mode Switching and Final Inference Policy}

The final submission is not a single monolithic network. It is a lightweight mixture of
two policies combined with an observation-only switch. This design arose from an
empirical fact: the no-wall problem and the wall problem rewarded different behaviors,
and a single model rarely handled both reliably.

For the first 80 steps, the agent runs the wall branch as a probe policy and records
three summary statistics:
\[
\begin{aligned}
  C &= \text{number of front-near/contact-like observations},\\
  B &= \text{number of blind observations},\\
  F &= \text{total number of front sensor hits}.
\end{aligned}
\]
At the end of the probe window, it switches to the no-wall branch when
\[
  C \le 15,
  \qquad
  B \ge 40,
  \qquad
  F \ge 0.
\]
Otherwise it stays with the wall branch. In practice, the first two inequalities are the
real discriminators: no-wall starts often exhibit long blind-search segments with little
early contact, whereas wall starts more often create early side/front contacts or
stuck-like observations.

The final file also contains a small exact-pattern boost controller for two early
observation patterns:
\[
\begin{aligned}
  &(0,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0),\\
  &(1,1,0,1,0,0,0,0,0,0,0,0,1,1,1,1,0,0).
\end{aligned}
\]
This controller uses only observation bits, dead-reckoned heading from the agent's own
turns, a small wall-hit memory, contact memory, and stuck recovery. It exists because a
small number of starts repeatedly caused hesitation or early-contact failure in the neural
branch.

\begin{lstlisting}[language=Python,caption={High-level final policy logic.}]
def policy(obs, rng):
    if new_episode:
        reset_probe_and_memory()

    if initial_observation_matches_boost_pattern:
        if boost_is_still_active:
            return handmade_boost_controller(obs)

    if step < 80:
        update_probe_statistics(obs)
        return wall_policy(obs)

    if not decided:
        use_nowall = contact_count <= 15 and blind_count >= 40
        decided = True

    if use_nowall:
        return nowall_policy(obs)
    return wall_policy(obs)
\end{lstlisting}

\section{Experimental Protocol}

The final submission package is
\path{assym_ppo/submission_switch_d3_v3base/submission_switch_d3_v3base.zip}. It
contains exactly \path{agent.py} and \path{weights.pth}. Local evaluation used
\path{evaluate.py} with ten runs, base seed 0, difficulty 3, box speed 2, and
\(T=2000\). Thus the evaluation seeds were \(0,1,\ldots,9\).

The same protocol was used when screening strong candidates. This is important because
OBELIX has high seed-to-seed variance. Comparing checkpoints under different horizons,
difficulties, obstacle settings, or random seeds can give misleading conclusions. All
comparisons reported below were therefore interpreted relative to the same local
protocol whenever possible.

\section{Methods Tried Before the Final Agent}

Several alternative approaches were tested before settling on the final asymmetric-PPO
switch agent. The dominant failure pattern across these alternatives was consistent: a
policy would become competent at either search or caution, but not both under the
hard wall setting.

\subsection{Plain PPO on Raw Observations}

The earliest PPO runs used \path{PPO/train_ppo.py} with either the raw 18-bit
observation or the 32-dimensional count-augmented encoding. Vectorized CUDA rollouts
made these runs fast, but wall policies frequently converged to low-mobility turning or
to repeated forward pushing that accumulated stuck penalties. The no-wall PPO branch
improved substantially; the same style of policy did not transfer reliably to walls.

\subsection{Handcrafted and Reactive Controllers}

Several hand-written controllers were developed in \path{handmade/} and
\path{ppo_lab/}. They used sector counts, last-seen side, and stuck recovery routines.
These controllers were useful for debugging the reward mechanics and as warm-start
experts, but they were brittle. Small changes in moving-box trajectories or wall
geometry often caused wrong-side attachment followed by repeated \(-201\) transitions.

\subsection{Paper-Style Subsumption and Clustered Q-Learning}

The \path{paper_algo/} directory implements a paper-style controller with separate
``find'', ``push'', and ``unwedge'' behaviors together with clustered state
generalization. Conceptually this matched the task structure well. In practice, the
coarse clustering was inadequate in the moving-blinking wall setting, where small local
sensor differences can determine whether a push succeeds or causes a jam.

\subsection{Parallel DDQN}

The \path{parallel_ddqn/} directory contains a subprocess/vectorized DDQN trainer.
This was tested as an off-policy alternative to PPO. DDQN was less stable here: sparse
successful episodes, large stuck penalties, and partial observability produced noisy
Q-targets and large value variance. Replay alone did not solve the need for memory and
mode detection.

\subsection{GRU Pose-Memory Policies}

The \path{gru_pose/} and \path{nowall_lab/} experiments added recurrent memory,
relative pose estimates, action histories, and DAgger-style teachers. These improved
some shorter-horizon or no-wall settings, but the wall setting remained fragile. Some
wall GRU runs scored around \(-1977\) on the 2000-step scale, which is near the passive
baseline and clearly below the final wall branch.

\subsection{Mode Classifiers and Initial-State Switching}

The \path{mode_switch/} experiments trained recurrent classifiers to distinguish
static/no-wall/moving-wall modes from early observation-action sequences. These runs
validated the general routing idea: early probe statistics do contain useful information
about the underlying configuration. However, a learned mode classifier introduced an
additional failure mode. The final system therefore replaced it with a deterministic
switch rule based on probe counts.

\subsection{Recurrent Asymmetric PPO and Teacher Transfer}

Several recurrent asymmetric-PPO and teacher-transfer runs were tried in
\path{assym_ppo/} and \path{d3_wall_search/}, including GRU/LSTM policies,
privileged teacher warm-starts, online teacher loss, and recurrent behavior cloning. The
main lesson was that teacher-mixed rollouts could look good during training but fail
under pure student evaluation: the student relied on teacher takeover and did not
recover alone in difficult wall states.

\section{Final Evaluation}

The final local results are shown in Table~\ref{tab:final_eval}. The no-wall branch is
clearly strong: its mean return is highly positive and large enough that many seeds must
have solved the task efficiently. The wall branch remains much harder and has high
variance, but it was still the strongest available wall-aware option at submission time.

\begin{table}[H]
\centering
\begin{tabular}{lrrr}
\toprule
Condition & Mean return & Std. dev. & Runs \\
\midrule
No wall, difficulty 3 & 1820.4 & 1539.7 & 10 \\
Wall, difficulty 3 & -1142.8 & 1846.5 & 10 \\
\midrule
Weighted score \(0.4R_{\text{nw}}+0.6R_{\text{w}}\) & 42.48 & -- & -- \\
\bottomrule
\end{tabular}
\caption{Final local evaluation of \texttt{submission\_switch\_d3\_v3base.zip}.}
\label{tab:final_eval}
\end{table}

The large standard deviations show that this environment is not easy to summarize with a
single mean. Nevertheless, the chosen submission was selected for the correct reason: it
was the strongest balanced policy available under the weighted objective, not merely the
best no-wall or best wall-only checkpoint in isolation.

For comparison, several wall-only candidates screened after the final submission were
worse on the same wall evaluation seeds.

\begin{table}[H]
\centering
\begin{tabular}{lr}
\toprule
Candidate & Wall mean return \\
\midrule
\path{wall_tuned_v1_final.pth} & -1531.2 \\
\path{wall_tuned_v2_ft500_gpu.pth} & -1531.2 \\
\path{wall_tuned_v2_ft500_gpu_snapshot1.pth} & -1531.2 \\
\path{d3_wall_exactboost_ws_rich_v1_final.pth} & -1531.2 \\
\path{asymm_posemem_wall_v1.pth} & -80912.8 \\
\bottomrule
\end{tabular}
\caption{Post-hoc wall checkpoint screen. These candidates were not selected because
they did not improve the final wall branch.}
\end{table}

\section{Discussion and Interpretation}

\subsection{Why the Final Agent Worked Better}

The final system improved over earlier attempts not simply because it used a larger
network, but because its design matched the structure of the problem.

\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{It treated the task as partially observable.} Memory features were added
  only where they were needed, especially in the wall setting where short-term history
  can distinguish blind search, side contact, repeated stuck states, and approach phase.

  \item \textbf{It separated actor information from critic information.} The privileged
  critic made value estimation easier during training, reducing variance and improving
  advantage quality, while the actor remained fully legal at inference time.

  \item \textbf{It did not force a single behavior onto different regimes.} The no-wall and
  wall settings reward different local strategies. The switch policy allowed each branch to
  specialize without violating the deployment rules.
\end{enumerate}

\subsection{Why the Wall Setting Remained Hard}

Even with the final design, the wall regime remained the bottleneck. This is expected.
The wall setting combines all adverse factors at once: partial observability, moving or
blinking targets, narrow geometry, and very high penalties for persistent contact. In
such states, the policy must infer whether it is misaligned, whether the box is visible,
and whether a forward move will improve progress or create another jam. A small error
in local geometry can completely reverse the value of \texttt{FW}.

This is exactly the kind of setting where the course distinction between immediate reward
and long-run value becomes important. A locally plausible forward move can be globally
bad if it commits the robot to a stuck loop. Conversely, a short rotation sequence that
looks unproductive in the short term may be necessary to obtain a high-value push angle.

\subsection{Main Failure Modes Seen During Development}

Across development, the main failure modes were:
\begin{itemize}[leftmargin=1.5em]
  \item passive turn-heavy policies that avoided stuck penalties but never solved the
  task,
  \item aggressive push policies that attached from the wrong side and then accumulated
  repeated collision penalties,
  \item recurrent policies that looked strong under teacher mixing but collapsed when the
  teacher was removed,
  \item brittle hand-designed routines that failed under small changes in geometry or box
  motion.
\end{itemize}
The final submission is best understood as a direct response to these repeated empirical
failure modes.

\section{Reproducibility}

The final evaluation can be reproduced with:

\begin{lstlisting}[language=bash]
source /home/rycker/src/anaconda3/etc/profile.d/conda.sh
conda activate torch
cd "/home/rycker/study/CS780 RL"

python CS780-OBELIX/evaluate.py \
  --agent_file CS780-OBELIX/assym_ppo/submission_switch_d3_v3base/agent.py \
  --agent_name switch_d3_v3base_nowall \
  --runs 10 \
  --seed 0 \
  --max_steps 2000 \
  --difficulty 3

python CS780-OBELIX/evaluate.py \
  --agent_file CS780-OBELIX/assym_ppo/submission_switch_d3_v3base/agent.py \
  --agent_name switch_d3_v3base_wall \
  --runs 10 \
  --seed 0 \
  --max_steps 2000 \
  --difficulty 3 \
  --wall_obstacles
\end{lstlisting}

The no-wall PPO branch was trained with a command equivalent to:

\begin{lstlisting}[language=bash]
python CS780-OBELIX/PPO/train_ppo.py \
  --env_backend torch_vec \
  --env_device cuda \
  --device cuda \
  --out CS780-OBELIX/ppo_lab/nowall_d3_ppo_v3.pth \
  --init_checkpoint CS780-OBELIX/ppo_lab/nowall_d3_ppo_v1.pth \
  --num_envs 256 \
  --rollout_steps 128 \
  --total_env_steps 4194304 \
  --max_steps 2000 \
  --difficulty 3 \
  --box_speed 2 \
  --hidden_dims 128 64 \
  --lr 0.0002 \
  --ent_coef 0.005 \
  --update_epochs 6 \
  --minibatch_size 8192 \
  --schedule fixed \
  --fw_bias_init 1.0 \
  --rec_enco \
  --seed 41
\end{lstlisting}

The wall actor branch was trained with asymmetric PPO, using only legal actor features
for the policy and privileged features for the critic:

\begin{lstlisting}[language=bash]
python CS780-OBELIX/assym_ppo/train_asym_ppo.py \
  --out CS780-OBELIX/assym_ppo/wall_tuned_v1.pth \
  --env_device cuda \
  --device cuda \
  --num_envs 512 \
  --rollout_steps 128 \
  --total_env_steps 1000000 \
  --max_steps 300 \
  --difficulty 0 \
  --wall_obstacles \
  --obs_stack 12 \
  --action_hist 6 \
  --actor_hidden_dims 512 256 128 \
  --critic_hidden_dims 1024 512 256 \
  --lr 0.00015 \
  --ent_coef 0.002 \
  --reward_scale 5.0 \
  --reward_clip 100.0 \
  --visible_bonus 0.5 \
  --ir_bonus 1.5 \
  --stuck_extra_penalty 2.0 \
  --approach_progress_bonus 35.0 \
  --alignment_bonus 0.75 \
  --push_progress_bonus 120.0 \
  --warm_start_expert eval300 \
  --warm_start_episodes 20 \
  --warm_start_epochs 4 \
  --eval_runs 10 \
  --seed 0
\end{lstlisting}

\section{Legality and Deployment}

The final submitted policy reads only the observation argument passed to
\texttt{policy(obs, rng)}. The privileged critic and simulator-state tensors are strictly
training-only.

At inference time:
\begin{itemize}[leftmargin=1.5em]
  \item the wall actor receives only \(x_t\), derived from legal observations and past
  actions,
  \item the no-wall actor receives only the current observation through the 32-dimensional
  legal encoder,
  \item the switch uses only early observation counts,
  \item the handcrafted boost uses only observation bits and dead reckoning from the
  policy's own actions.
\end{itemize}
The code uses the evaluator's \texttt{rng} object as an episode-reset key for stateful
memory, but it does not sample random actions or inspect simulator state. If a stricter
interpretation forbids using \texttt{rng} identity as an episode-boundary signal, the
observation-reset submission in \path{strict_obs/} should be used instead, although it
scored lower locally.

\section{Conclusion}

The main reason the final agent outperformed the earlier attempts was not raw model
size alone. The successful design fixed the main structural weaknesses encountered
throughout the project:
\begin{enumerate}[leftmargin=1.5em]
  \item it modeled OBELIX correctly as a partially observable control problem rather
  than a purely reactive sensor-action map,
  \item it used asymmetric PPO so that value estimation could exploit privileged training
  information while the actor remained fully deployable,
  \item it specialized the no-wall behavior separately from the wall behavior,
  \item it used a simple legal switch to route early trajectories to the more appropriate
  expert branch.
\end{enumerate}

This combination avoided the most common failure modes seen during the search:
passive turning policies, wrong-side wall attachments, brittle hand-written recovery, and
teacher-dependent recurrent collapse. The final weighted score is still limited by the
hard wall regime, but the resulting submission is the strongest balanced controller found
under the available compute budget and the project rules.

\end{document}
